---
title: "Milestone 4"
author: "Wyatt Hurt"
date: "03/06/2020"
output: html_document
---
[Link to GitHub repository](https://github.com/wyatthurt/gov1005-final-project).
<br><br>

This week, I began manipulating my data in ways that allow me to derive useful insights from it. I focused on creating "proof of concepts," where I figure out how to use R to manipulate my data for statistical inference and transform it into visualizations. As I continue, I will be able to replicate the architectures of these "proof of concepts" and apply them to different components of my project. 

## Mapping 
I am very interested in spatial data and am hoping to incorporate maps into my project. I began by downloading river basin shapefiles from the World Bank. It took me several different tries (and packages) before I figured out that readOGR(), from the rgdal package, was the best option, despite the fact that it produces a strange output: a PolygonsDataFrame. Wrestling with this strange format became more important when creating my demo map.
```{r setup_data, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# I turned off warnings and messages temporarily for the purpose of knitting
# this document. Since I am actively in the process of manipulating data, I have
# not been able to eliminate the sources of all these warnings and messages.

# Running a script to gather my data (slightly modified from milestone 3):

library(rmarkdown)
render("gather.R")

# Importing World Bank shapefiles for the world's major river basins ( https://datacatalog-worldbank-org/dataset/major-river-basins-world):  

basins_geometry <- readOGR("geometry/wb_major_basins", layer = "Major_Basins_of_the_World") 
```

Initially, I used the get_map() function and the ggmap package to create a static map of the world's river basins. However, I was unhappy with this approach because it produced a low-resolution product, and it didn't allow my audience to interact with the map. I used the leaflet library to resolve both of these problems. I decided to create a demo plot of the world's river basins, color filled by their relative level of water conflict since 1948 (when the OSU dataset begins). 
```{r mapping}

# Created a tibble with a list of events. I chose to filter for strings that
# include words related to conflict. I need to refine my filtering approach, but
# using these four terms works as an initial proxy for my demo case. To be able
# to merge this data with my polygons, I produce a per-basin count of
# conflict-related events.

events_n <- joined %>%
  filter(str_detect(event_summary, c("conflict", "war", "violence", "military"))) %>%
  distinct(date, .keep_all = TRUE) %>%
  group_by(basin_name) %>%
  count() %>%
  rename(num_events = n)

# I then merge these counts with my polygon data. Leaflet throws an error if
# the variable column includes NA values, so I replaced all NA values.

basins_geometry@data <- merge(basins_geometry@data, events_n, by.x = "NAME", by.y = "basin_name") %>%
  replace_na(list(num_events = 0))

# Defining my leaflet color bin:

binpal <- colorBin("Blues", basins_geometry$num_events, 9, pretty = FALSE)

# Created my Leaflet map, using a simple CartoDB basemap. Zoomed out and
# centered the map, added my polygons, and changed their color based on their
# relative number of events. Added a legend.

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(lng = 0, lat=30,zoom=1.3) %>% 
  addPolygons(data = basins_geometry, stroke = FALSE, smoothFactor = 0.2, 
              fillOpacity = .6, color = ~binpal(num_events)) %>%
    addLegend("bottomright", pal = binpal, values = basins_geometry$num_events, 
              title = "# of water conflict events since 1948", opacity = 1)

# To do: Need to figure out how to add static or hover labels to the river basin
# polygons.
```

## Case Study: Jordan River Basin
I plan to present both aggregate data and specific case studies in my project. To check data quality and test my ability to manipulate the OSU data to produce basin-specific insights, I filtered for water conflict events in the Jordan River Basin.
```{r jor_data}
jord <- joined %>%
  filter(bcode == "JORD") %>%
  filter(str_detect(event_summary, c("conflict", "war", "violence", "military"))) %>%
  arrange(date) %>%
  distinct(date, .keep_all = TRUE)
```

Then, I created a frequency histogram of these events over time. The relatively calm period in the 1970s and 80s is consistent with a period of stagnation that I've written about previously for other research projects.
```{r jor_hist}
ggplot(jord, aes(year)) + 
  geom_histogram(bins = 60) + 
  labs(title = "Water Conflict Events in the Jordan River Basin, 1948-2008", 
       caption = "Source: OSU Program in Water Conflict Management and Transformation") + 
  xlab("Year") + ylab("Count") + 
  theme_classic()
```
<br><br>
Next, I created a table that displays the water conflict events in the basin. After looking at the events being returned, it seems like a more accurate title might be "Water Conflict and Diplomacy Events", since a number of the events in the OSU database are not so much violent conflict as violent rhetoric and/or negotiations to avoid conflict. 
I have conducted research on the Jordan River Basin for over a year now; in my opinion, these events appropriately characterize the water conflict events that occured over this period. This reflects positively on the quality of the OSU dataset.
```{r jor_table}
jord_gt <- jord %>%
  select(date, dyad_code, event_summary)

gt(jord_gt) %>%
  tab_header("Water Conflict Events in the Jordan River Basin, 1948-2008") %>%
  cols_label(date = "Date", 
             dyad_code = "Conflict Dyad", 
             event_summary = "Event Summary")
```
<br><br>

## Future Plans
Next week, I hope to expand the depth and breadth of my analysis of the Jordan River Basin and develop case studies of the Mekong, Indus, and Colorado river basins. I chose these basins in part based on insights derived from the rough Leaflet map in this milestone: Based on my background knowledge of transboundary river basins and the map, I selected cases that are roughly comparable demographically and geopolitically, but which have wide variability in the number of water conflicts they have experienced.
